\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-01-29 src/010chap1.tex
- a whole bunch of cleanups. see diff. - removed problem defs from chap 1. they were duplicates from chap 3. - CODED FIGURES IN TIKZ. forbidden subgraphs, graphmatrix example graphs. - BEGAN CHAPTER 3 EDITS.
Formatting/layout changes
-------

- Added FrenchLayout to make lists denser
- Added hanging indent for Definitions
- Compressed theorem env's above below spaces
- Creaded env for problem definition minipages.
- moved all clearpage/page counter and other prechapter settings to the chapter tex files.
- Fixed the page number bug in draft printout. removed \fancyhf{} to avoid resetting page n

\subsection{Special case}
\label{sec:courseschedule}
As a special case of the study group accommodation problem, suppose
all the apartments are on the same street or if they are all lined up
on a single path, the street map becomes a tree that is just a
path. Then the problem becomes what is called an {\em interval
  assignment problem}. The idea of interval assignment may not be
obvious here; hence to see this, consider a different problem in
{\WSI} where the classes for these study groups courses need to be
scheduled during a day (or a week or any time period). Each study
group has a bunch of courses associated with it some of which may be
shared by two or more study groups. It is mandatory that a student who
is a member of a study group takes all the courses associated with
that group. There are slots during the day for classes to be held and
the problem is to allocate class slots to courses such that all the
classes of a study group are consecutive. It is debatable if this will
not hamper the attention span and memory retention rate of the
students but that is, regrettably, out of the scope of this
thesis. The parallels between this class allocation problem and the
accommodation problem can be seen as follows. The set $U$ here, are
the courses offered (say Course 101 {\coneohone}, Course 102
{\coneohtwo} and so on). In this variation of the problem, the
collection $\cF$ is the set of study groups but the study groups are
filled by course IDs (in place of students in the earlier
example). For instance, Course 101 is mandatory for all study groups
$\xLLL$, $\xGGG$, $\xBBB$, $\xTTT$ and Course 102 is mandatory for
only the $\xLLL$ group) and so on. The sequence of class slots for the
day (or week or any time period) is analogous to the street map in the
accommodation problem. It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree) is called an interval assignment
problem.

@@

\begin{definition}
  A {\em permutation} $\lambda$ of a set $X = \{x_1, x_2, \ldots
  x_n\}$ is a bijection $\lambda: X \rightarrow X$.  For simplicity,
  sometimes, $\lambda$ is written as a sequence $x_{i_1}x_{i_2}\ldots
  x_{i_n}$, where $i_j \in \{1, 2 \ldots n\}$ to mean that
  $\lambda(x_j)= x_{i_j}$.\\
  This document uses both notations as convenient in context.
\end{definition}

@@

\begin{definition}[\emph{Graph, Tree}]
  \label{def:graphtree}  
\end{definition}

@@

We see in Section~\ref{sec:motive} that pairwise intersection
cardinality preservation is necessary and sufficient for an interval
assignment to be feasible for a given hypergraph\footnote{A {\em
    hypergraph} is an alternate representation of a set system and
  will be used in this thesis.}~\footnote{See Section~\ref{ch:prelims}
  for the formal definition.} and thus is a characterization for COP
\cite{nsnrs09}. In our work we extend this characterization and find
that trio-wise intersection cardinality preservation makes a tree path
labeling\footnote{A {\em tree path labeling} $\cl$ is a bijection of
  paths from the target tree $T$ to the hyperedges in given hypergraph
  $\cF$.}~\footnotemark[4] (TPL) feasible, which is a generalization
of the COP problem. This problem is defined as follows.

{\small
  \begin{minipage}[h]{5in}
    % \begin{singlespace}
    \vspace{2mm}
    {\large \FTPL}\\
    \begin{tabular}[t]{l|l}
      \hline\\
      {\tt Input} & 
      \begin{minipage}[t]{\probdefwidth}
        A hypergraph $\cF$ with vertex set $U$, a tree $T$, a set of
        paths $\cP$ from $T$ and a
        bijection $\cl$~$:$~$\cF \rightarrow \cP$.\\
      \end{minipage}\\
      {\tt Question} &
      \begin{minipage}[t]{\probdefwidth}
        Does there exist a bijection $\phi$~$:$~$U \rightarrow V(T)$
        such that $\phi$ when applied on any hyperedge in $\cF$ will
        give
        the path mapped to it by the given tree path labeling $\cl$.\\
        { i.e., $\cl(S) = \{\phi(x) \mid x \in S\}$, for every
          hyperedge $S \in \cF$.}
      \end{minipage}\\
    \end{tabular}
    % \end{singlespace}
  \end{minipage}\\
}

@@

Aside from checking if a given TPL is feasible, we also solve the
problem of computing a feasible TPL for a given hypergraph and target
tree, if one exists. This problem, {\CFTPL}, is defined as follows.

{\small
  \begin{minipage}[h]{5in}
    % \begin{singlespace}
    \vspace{2mm}
    {\large \CFTPL}\\
    \begin{tabular}[t]{l|l}
      \hline\\
      {\tt Input} & 
      \begin{minipage}[t]{\probdefwidth}
        A hypergraph $\cF$ with vertex set $U$ and a tree $T$.\\
      \end{minipage}\\

      {\tt Question} &
      \begin{minipage}[t]{\probdefwidth}
        Does there exist a set of paths $\cP$ from $T$ and a bijection
        $\cl$~$:$~$\cF \rightarrow \cP$, such that {\FTPL} returns
        {\bf true} on $(\cF, T, \cl)$.
      \end{minipage}\\
    \end{tabular}
    % \end{singlespace}
  \end{minipage}\\
}

We present a polynomial time algorithm for {\CFTPL} when the target
tree $T$ belongs to a special class of trees called {\em \kstars} and
when the hyperedges in the hypergraph $\cF$ have at most $k+2$
vertices. A couple of examples of {\kstars} are given in
Figure~\ref{fig:ksubstar}.

@@

{\small
  \begin{minipage}[h]{5in}
    % \begin{singlespace}
    \vspace{2mm}
    {\large \CFTPLKTREE}\\
    \begin{tabular}[t]{l|l}
      \hline\\
      {\tt Input} & 
      \begin{minipage}[t]{\probdefwidth}
        A hypergraph $\cF$ with vertex set $U$ such that every
        hyperedge
        $S \in \cF$ is of cardinality at most $k+2$ and a {\kstar} $T$.\\
      \end{minipage}\\
      {\tt Question} &
      \begin{minipage}[t]{\probdefwidth}
        Does there exist a set of paths $\cP$ from $T$ and a bijection
        $\cl$~$:$~$\cF \rightarrow \cP$, such that {\FTPL} returns
        {\bf true} on $(\cF, T, \cl)$.
      \end{minipage}\\
    \end{tabular}
    % \end{singlespace}
  \end{minipage}\\
}

\temptext{The following sentence seems out of place in this para} In
spite of this being a restricted case, we believe that our results are
of significant interest in understanding the nature of {\sc Graph
  Isomorphism} which is polynomial time solvable in interval graphs
while being hard on path graphs\cite{kklv10}. {\kstars} are a class
of trees which are in many ways very close to intervals or paths. Each
ray\footnote{The path from a leaf to the root, the vertex with highest
  degree, is called a {\em ray} of the \kstar.}~\footnotemark[4] are independent except for the
root\footnote{The vertex with maximum degree in a {\kstar} is called
  {\em root}.}~\footnotemark[4] and hence can be
considered as an independent interval till the root. Our algorithm
builds on this fact and uses the interval assignment
algorithm\cite{nsnrs09} up until ``reaching'' the root and then uses
the trio-wise intersection cardinality (the extra condition in ICPPL
that generalizes ICPIA) check to resolve the ambiguity about which ray
the algorithm should ``grow'' the solution into in the next iteration.

>>>>

\subsection{Special case}
\label{sec:courseschedule}
As a special case of the study group accommodation problem, suppose
all the apartments are on the same street or if they are all lined up
on a single path, the street map becomes a tree that is just a
path. Then the problem becomes what is called an {\em interval
  assignment problem}. The idea of interval assignment may not be
obvious here; hence to see this, consider a different problem in
{\WSI} where the classes for these study groups courses need to be
scheduled during a day (or a week or any time period). Each study
group has a bunch of courses associated with it some of which may be
shared by two or more study groups. It is mandatory that a student who
is a member of a study group takes all the courses associated with
that group. There are slots during the day for classes to be held and
the problem is to allocate class slots to courses such that all the
classes of a study group are consecutive. % \footnote{It is debatable if this will
% not hamper the attention span and memory retention rate of the
% students but that is undoubtedly out of the scope of this
% thesis.}
The parallels between this class allocation problem and the
accommodation problem can be seen as follows. The set $U$ here, are
the courses offered (say Course 101 {\coneohone}, Course 102
{\coneohtwo} and so on). In this variation of the problem, the
collection $\cF$ is the set of study groups but the study groups are
filled by course IDs (in place of students in the earlier
example). For instance, Course 101 is mandatory for all study groups
$\xLLL$, $\xGGG$, $\xBBB$, $\xTTT$ and Course 102 is mandatory for
only the $\xLLL$ group) and so on. The sequence of class slots for the
day (or week or any time period) is analogous to the street map in the
accommodation problem. It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree) is called an interval assignment
problem.

@@

\begin{definition}
  A {\em permutation} $\lambda$ of a set $X = \{x_1, x_2, \ldots
  x_n\}$ is a bijection $\lambda: X \rightarrow X$.  For simplicity,
  $\lambda$ may be written as a sequence $x_{i_1}x_{i_2}\ldots x_{i_n}$,
  where $i_j \in [n]$ to mean that $\lambda(x_j)= x_{i_j}$.
  This document uses both notations as convenient in context.
\end{definition}

@@

\begin{definition}[\emph{Graph, Tree}]
  \label{def:graphtree}

\end{definition}
When refering to a tree as $T$ it could be a reference to the tree
itself, or the vertices of the tree. This will be clear from the
context.

@@

We see in Section~\ref{sec:motive} that pairwise intersection
cardinality preservation is necessary and sufficient for an interval
assignment to be feasible for a given hypergraph\footnote{A {\em
    hypergraph} is an alternate representation of a set system and
  will be used in this thesis.}~\footnote{See Section~\ref{ch:prelims}
  for the formal definition.} and thus is a characterization for COP
\cite{nsnrs09}. In our work we extend this characterization and find
that trio-wise intersection cardinality preservation makes a tree path
labeling\footnote{A {\em tree path labeling} $\cl$ is a bijection of
  paths from the target tree $T$ to the hyperedges in given hypergraph
  $\cF$.}~\footnotemark[4] (TPL) feasible, which is a generalization
of the COP problem. This problem is defined by \FTPL (See
Section~\ref{sec:myresearchintro} for problem definition). %as follows.

% {\small
%   \begin{minipage}[h]{5in}
%     % \begin{singlespace}
%     \vspace{2mm}
%     {\large \FTPL}\\
%     \begin{tabular}[t]{l|l}
%       \hline\\
%       {\tt Input} & 
%       \begin{minipage}[t]{\probdefwidth}
%         A hypergraph $\cF$ with vertex set $U$, a tree $T$, a set of
%         paths $\cP$ from $T$ and a
%         bijection $\cl$~$:$~$\cF \rightarrow \cP$.\\
%       \end{minipage}\\
%       {\tt Question} &
%       \begin{minipage}[t]{\probdefwidth}
%         Does there exist a bijection $\phi$~$:$~$U \rightarrow V(T)$
%         such that $\phi$ when applied on any hyperedge in $\cF$ will
%         give
%         the path mapped to it by the given tree path labeling $\cl$.\\
%         { i.e., $\cl(S) = \{\phi(x) \mid x \in S\}$, for every
%           hyperedge $S \in \cF$.}
%       \end{minipage}\\
%     \end{tabular}
%     % \end{singlespace}
%   \end{minipage}\\
% }

@@

Aside from checking if a given TPL is feasible, we also solve the
problem of computing a feasible TPL for a given hypergraph and target
tree, if one exists. This problem is the {\CFTPL} problem (See
Section~\ref{sec:myresearchintro} for problem definition). %, is defined as follows.

% {\small
%   \begin{minipage}[h]{5in}
%     % \begin{singlespace}
%     \vspace{2mm}
%     {\large \CFTPL}\\
%     \begin{tabular}[t]{l|l}
%       \hline\\
%       {\tt Input} & 
%       \begin{minipage}[t]{\probdefwidth}
%         A hypergraph $\cF$ with vertex set $U$ and a tree $T$.\\
%       \end{minipage}\\

%       {\tt Question} &
%       \begin{minipage}[t]{\probdefwidth}
%         Does there exist a set of paths $\cP$ from $T$ and a bijection
%         $\cl$~$:$~$\cF \rightarrow \cP$, such that {\FTPL} returns
%         {\bf true} on $(\cF, T, \cl)$.
%       \end{minipage}\\
%     \end{tabular}
%     % \end{singlespace}
%   \end{minipage}\\
% }

We present a polynomial time algorithm for {\CFTPL} when the target
tree $T$ belongs to a special class of trees called {\em \kstars} and
when the hyperedges in the hypergraph $\cF$ have at most $k+2$
vertices (\CFTPLKTREE -- See Section~\ref{sec:myresearchintro} for
problem definition). A couple of examples of {\kstars} are given in
Figure~\ref{fig:ksubstar}.

@@

% {\small
%   \begin{minipage}[h]{5in}
%     % \begin{singlespace}
%     \vspace{2mm}
%     {\large \CFTPLKTREE}\\
%     \begin{tabular}[t]{l|l}
%       \hline\\
%       {\tt Input} & 
%       \begin{minipage}[t]{\probdefwidth}
%         A hypergraph $\cF$ with vertex set $U$ such that every
%         hyperedge
%         $S \in \cF$ is of cardinality at most $k+2$ and a {\kstar} $T$.\\
%       \end{minipage}\\
%       {\tt Question} &
%       \begin{minipage}[t]{\probdefwidth}
%         Does there exist a set of paths $\cP$ from $T$ and a bijection
%         $\cl$~$:$~$\cF \rightarrow \cP$, such that {\FTPL} returns
%         {\bf true} on $(\cF, T, \cl)$.
%       \end{minipage}\\
%     \end{tabular}
%     % \end{singlespace}
%   \end{minipage}\\
% }

In spite of this being a restricted case, we believe that our results
are of significant interest in understanding the nature of {\sc Graph
  Isomorphism} which is polynomial time solvable in interval graphs
while being hard on path graphs\cite{kklv10}. {\kstars} are a class of
trees which are in many ways very close to intervals or paths. Each
ray\footnote{The path from a leaf to the root, the vertex with highest
  degree, is called a {\em ray} of the \kstar.}~\footnotemark[4] are
independent except for the root\footnote{The vertex with maximum
  degree in a {\kstar} is called {\em root}.}~\footnotemark[4] and
hence can be considered as an independent interval till the root. Our
algorithm builds on this fact and uses the interval assignment
algorithm\cite{nsnrs09} up until ``reaching'' the root and then uses
the trio-wise intersection cardinality (the extra condition in ICPPL
that generalizes ICPIA) check to resolve the ambiguity about which ray
the algorithm should ``grow'' the solution into in the next iteration.

### 2012-01-29 src/020chap2.tex
- a whole bunch of cleanups. see diff. - removed problem defs from chap 1. they were duplicates from chap 3. - CODED FIGURES IN TIKZ. forbidden subgraphs, graphmatrix example graphs. - BEGAN CHAPTER 3 EDITS.
Formatting/layout changes
-------

- Added FrenchLayout to make lists denser
- Added hanging indent for Definitions
- Compressed theorem env's above below spaces
- Creaded env for problem definition minipages.
- moved all clearpage/page counter and other prechapter settings to the chapter tex files.
- Fixed the page number bug in draft printout. removed \fancyhf{} to avoid resetting page n

\definecolor{gray}{rgb}{0.8,0.8,0.8}
\def\tblhcolor {gray}

@@

\temptext{ADD: have a few lines about organization of chapter}

@@

\COP is closely connected to several types of graphs by way of
describing certain combinatorial graph properties. There are also
certain graphs, like convex bipartite graphs, that are defined solely
by some of its associated matrix having \COP.  In this section we will
see the relevance of \cop to graphs.  To see this we introduce certain
binary matrices that are used to define graphs in different
ways. While adjacency matrix is perhaps the most commonly used such
matrix, Definition~\ref{def:graphmatrices} defines this and a few
more.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.35]{../img/tuckersforbiddensubgraphs.png}  
  \caption[\figtabsize Tucker's forbidden subgraphs]{\figtabsize Tucker's
    forbidden subgraphs for convex bipartite graphs. \temptext{ PLACEHOLDER
    IMG }} 
  \label{fig:forbiddensubgraphs}
\end{figure}

@@

After the invention of \PQtrees, presumably due to the implementation
challenge it posed, there has been several variants of the same in the
literature, like \PCtree \cite{sh99,wlh01,hm03}, generalized \PQtree
\cite{km89,mcc04}, \PQRtree \cite{mm96,mpt98} etc. Most of these are
generalizations of \PQtree -- for instance, \PCtree is generalized to
matrices with \CROP, \PQRtree and generalized \PQtree are generalized
to matrices and set systems with or without \COP. \cite{km89} invented
a modified form of \PQtree a simpler incremental update of the tree
only for recognizing interval graphs.  \cite{kr88} constructed
efficient parallel algorithms for manipulating PQ trees. \temptext{dom
  chapter 2 pg 40 Variations of PQTrees first para - summarize.}

In the next few following sections we will see some of these
variations.
%Their algorithm was improved by \cite{tm05}
%\tnote{improved in terms of what?}. 
% describes an improved algorithm to
% build PQR trees.

@@

\begin{definition}
  \label{def:pctree}
  [\emph{\PCtree} \cite{wlh01,d08phd}.]  A \PCtree of matrix $M$ with
  \CROP on columns (rows), is a tree with the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Is unrooted -- thus it has (a) no parent child relationship
    between nodes (b) there is no left to right (or vice versa)
    ordering.
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \CROP order for column (row) for
    $M$. Moreover, any sequence obtained by considering the leaves in
    clockwise or counter-clockwise order describes a \CROP order for
    $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $C$.
  \item The neighbors of $P$ nodes can be permuted in any fashion to
    obtain a new \CROP order for $M$.
  \item \label{def::pcmirror} The tree can be changed by applying the following
    ``mirroring'' operation to obtain a new \CROP order for $M$. Root
    the \PCtree at a neighbor of a $C$-node, $v$ and mirror the
    subtree whose root is $v$ and finally unrooting the
    tree. Mirroring a subtree is done by putting the children of every
    node of the subtree in reverse order.
  \end{enumerate}
\end{definition}

@@

\subsection{ICPIA - a set cardinality based \COP test}

\temptext{narrative is abrupt}

>>>>

\xclearpage

@@

\tnote[dire]{ADD: have a few lines about organization of chapter}

@@

\COP is closely connected to several types of graphs by way of
describing certain combinatorial graph properties. There are also
certain graphs, like convex bipartite graphs, that are defined solely
by one of its associated matrix having \COP.  In this section we will
see the relevance of \cop to graphs.  To see this we introduce certain
binary matrices that are used to define graphs in different
ways. While adjacency matrix is perhaps the most commonly used such
matrix, Definition~\ref{def:graphmatrices} defines this and a few
more.

@@

\figforbiddensubgraphs

@@

After the invention of \PQtrees, presumably due to the implementation
challenge it posed, there has been several variants of the same in the
literature, like \PCtree \cite{sh99,wlh01,hm03}, generalized \PQtree
\cite{km89,mcc04}, \PQRtree \cite{mm96,mpt98} etc. Most of these are
generalizations of \PQtree\ -- for instance, \PCtree is generalized to
matrices with \CROP, \PQRtree and generalized \PQtree are generalized
to matrices and set systems with or without \COP. \cite{km89} invented
a modified form of \PQtree a simpler incremental update of the tree
only for recognizing interval graphs.  \cite{kr88} constructed
efficient parallel algorithms for manipulating \PQtrees. \temptext{dom
  chapter 2 pg 40 Variations of PQTrees first para - summarize.}

In the next few sections we will see some of these
variations.
%Their algorithm was improved by \cite{tm05}
%\tnote{improved in terms of what?}. 
% describes an improved algorithm to
% build PQR trees.

@@

\begin{definition}[\PCtree \cite{wlh01,d08phd}]
  \label{def:pctree}
   A \PCtree of matrix $M$ with
  \CROP on columns (rows), is a tree with the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Is unrooted -- thus it has (a) no parent child relationship
    between nodes (b) there is no left to right (or vice versa)
    ordering.
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \CROP order for column (row) for
    $M$. Moreover, any sequence obtained by considering the leaves in
    clockwise or counter-clockwise order describes a \CROP order for
    $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $C$.
  \item The neighbors of $P$ nodes can be permuted in any fashion to
    obtain a new \CROP order for $M$.
  \item \label{def::pcmirror} The tree can be changed by applying the following
    ``mirroring'' operation to obtain a new \CROP order for $M$. Root
    the \PCtree at a neighbor of a $C$-node, $v$ and mirror the
    subtree whose root is $v$ and finally unrooting the
    tree. Mirroring a subtree is done by putting the children of every
    node of the subtree in reverse order.
  \end{enumerate}
\end{definition}

@@

\subsection{ICPIA - a set cardinality based \COP test}
\label{sec:icpiasurvey}
\temptext{narrative is abrupt}
